Adding selection to \stdpopsim: oh how the tables have turned

### 2024-12-11 analysis2_manuscript.tex
edits throughout

@@

    % DRS: This figure could really use some Python API and comand line examples like in Figs 1B,C
    % form the original stdpopsim paper
    \begin{wrapfigure}[37]{l}{0.4\textwidth}
        \vspace{-0.0cm}
        \includegraphics[width=\linewidth]{figures/schematics/catalog.pdf}
        \caption{\label{fig:schematic}
        A schematic of the \stdpopsim catalog and the different components that
        can be used in simulations. The user specifies a species, a portion of
        the genome to simulate, and optionally a genetic map, a model of
        demography, and a model of selection that itself is composed of a
        distribution of fitness effects (DFE) and a set of functional
        annotations.}
    \end{wrapfigure}
    \hfill

@@

    Using \stdpopsim, we can easily simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography}, we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without purifying selection on exons. For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome.
    % DRS: Maybe we should say something about how easy it was to do these simulations, and maybe even show
    % some example code somewhere to prove our point? Could help emphasize the utility of this resource.
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted computational resources as thorough benchmarking
    is not the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography}, we show six panels, each of which shows the estimated $N_e(t)$
    from data simulated under the five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The top row shows the results of $N_e(t)$ estimates made from simulations without selection, while the bottom row shows the results
    from simulations with deleterious mutations occurring in exons.
    In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the top row are to visually reinforce that inference is done
    from a model without selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ for the CHB sample and to a lesser extent in the CEU sample. #ADK: is GONE time scaled incorrectly? i.e. gens not years?
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwayplot, which uses the SFS to infer $N_e(t)$, 
    is reasonably close to the true $N_e(t)$ both with and without selection,
    although it seems to infer significant population size changes that are not part of the true history.
    Likewise \smcpp seems to be robust to the presence of selection,
    but the timing of population size changes is slightly inaccurate. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$ particularly in the very recent past,
    although estimates from \msmc have quite a bit of added noise even in the absence of selection.

@@

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without purifying selection on exons. The top row shows estimates of $N_e(t)$ from simulations
        without selection, while the bottom row shows estimates of $N_e(t)$ from simulations with a gamma-distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        }
    \end{figure}
    %DRS-fig: reminder to update the legend to just say ``exons'' instead of ensembl_havana_whatever...

@@

    Using this set of simulations, we can also compare the power of different methods to detect sweeps.
    In general, the power of the methods varies with respect to the population history of the sample.
    For instance \sweepfinder2 has marginally higher power to detect sweeps in the CEU sample than \diploshic,
    likely as a results of the fact that we here trained \diploshic on simulations under a constant size population,
    while the population size history of CEU is characterized by a strong recent bottleneck.
    Conversely, \diploshic has higher power to detect sweeps in the YRI sample than \sweepfinder2,
    likely because that population size history more closely matches the population size history used to train \diploshic.
    \sweepfinder2 power remains more constant along the chromosome, while the $\pi$ cutoff method varies wildly.

    We can further explore the performance of the methods by plotting the receiver operating characteristic (ROC) curves for the methods.
    In Figure \ref{fig:roc-curves}, we show the ROC curves for the three methods for detecting sweeps.
    False positive rates were established using neutral simulations that contained no sweeps
    but were conditioned on the same population history as the CEU and YRI samples.
    These suggest that the $\pi$ cutoff is a suprisingly effective method for detecting sweeps,
    in these genomic regions at least, but that it lags behind \diploshic and \sweepfinder2 in terms of accuracy in the YRI sample.
    %While this is not exhaustive benchmarking, we aim to give a sense to the reader
    %of how \stdpopsim can be used to benchmark the performance of different methods easily and efficiently.
    These ROC curves, however, hide the phenomenon of high variation in power (see Figure \ref{fig:chr1_power}) to detect sweeps along a realistic looking chromosome.
    That is, although in aggregate the $\pi$ cutoff may seem effective, it's power is highly heterogeneous across the chromosome.

@@

    To further dissect the reasons for high variance (along the chromosome) in power to detect sweeps,
    we show in Figure \ref{fig:power-recomb} the relationship between power and local recombination rate for our sweep simulations.
    We see that power to detect sweeps is
    a decreasing function of local recombination rate for \sweepfinder2 and $pi$ outliers,
    which is consistent with sweeps having a larger genomic footprint in regions of low
    recombination.
    \diploshic on the other hand shows a slight increase in power with recombination rates,
    probably because it is designed to distinguish regions where a sweep happend from regions linked to a sweep.
    We also see that the presence of purifying selection acting on exons has a relatively small effect on the power to detect sweeps,    
    but one that is present across the entire range of recombinations rates observed in human chromosome 1. 
    Unsurprisingly, the empirical cutoff of $\pi$ to detect sweeps is the less robust to the effects of recombination rate than the other two methods,
    which are specifically designed to find selective sweeps.

>>>>

    \begin{wrapfigure}{l}{0.5\textwidth}
        \vspace{-0.0cm}
        \includegraphics[width=\linewidth]{figures/schematics/catalog.pdf}
        \caption{\label{fig:schematic}
        \textbf{(A)} A schematic of the \stdpopsim catalog and the different components that
        can be used in simulations. 
        \textbf{(B)} The python API and \textbf{(C)} the command line interface.
        The user specifies a species, a portion of
        the genome to simulate, and optionally a genetic map, a model of
        demography, and a model of selection that itself is composed of a
        distribution of fitness effects (DFE) and a set of functional
        annotations.}
    \end{wrapfigure}
    \hfill

@@

@@

    Using \stdpopsim, we can easily simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography}, we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without purifying selection acting, either on exons or on every site in the genome.
    For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome.
    % DRS: Maybe we should say something about how easy it was to do these simulations, and maybe even show
    % some example code somewhere to prove our point? Could help emphasize the utility of this resource.
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted computational resources as thorough benchmarking
    is not the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography}, we show six panels, each of which shows the estimated $N_e(t)$
    from data simulated under the five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The left column shows the results of $N_e(t)$ estimates made from simulations without selection, 
    while the right column shows the results
    from simulations with deleterious mutations occurring in exons.
    In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the top row are to visually reinforce that inference is done
    from a model without selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ for the CHB sample and to a lesser extent in the CEU sample.
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwayplot, which uses the SFS to infer $N_e(t)$, 
    is reasonably close to the true $N_e(t)$ both with and without selection,
    although it seems to infer significant population size changes that are not part of the true history.
    Likewise \smcpp seems to be robust to the presence of selection,
    but the timing of population size changes is slightly inaccurate. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$ particularly in the very recent past,
    although estimates from \msmc have quite a bit of added noise even in the absence of selection.

@@

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without purifying selection on exons. The left column shows estimates of $N_e(t)$ from simulations
        without selection, while the right column shows estimates of $N_e(t)$ from simulations with a gamma-distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        }
    \end{figure}
    %DRS-fig: reminder to update the legend to just say ``exons'' instead of ensembl_havana_whatever...

@@

    Using this set of simulations, we can also compare the power of different methods to detect sweeps.
    In general, the power of the methods varies with respect to the population history of the sample.
    For instance \sweepfinder has marginally higher power to detect sweeps in the CEU sample than \diploshic,
    likely as a results of the fact that we here trained \diploshic on simulations under a constant size population,
    while the population size history of CEU is characterized by a strong recent bottleneck.
    Conversely, \diploshic has higher power to detect sweeps in the YRI sample than \sweepfinder,
    likely because that population size history more closely matches the population size history used to train \diploshic.
    \sweepfinder power remains more constant along the chromosome, while the $\pi$ cutoff method varies wildly.

    We can further explore the performance of the methods by plotting the receiver operating characteristic (ROC) curves for the methods.
    In Figure \ref{fig:roc-curves}, we show the ROC curves for the three methods for detecting sweeps.
    False positive rates were established using neutral simulations that contained no sweeps
    but were conditioned on the same population history as the CEU and YRI samples.
    These suggest that the $\pi$ cutoff is a suprisingly effective method for detecting sweeps,
    in these genomic regions at least, but that it lags behind \diploshic and \sweepfinder in terms of accuracy in the YRI sample.
    %While this is not exhaustive benchmarking, we aim to give a sense to the reader
    %of how \stdpopsim can be used to benchmark the performance of different methods easily and efficiently.
    These ROC curves, however, hide the phenomenon of high variation in power (see Figure \ref{fig:chr1_power}) to detect sweeps along a realistic looking chromosome.
    That is, although in aggregate the $\pi$ cutoff may seem effective, it's power is highly heterogeneous across the chromosome.

@@

    To further dissect the reasons for high variance (along the chromosome) in power to detect sweeps,
    we show in Figure \ref{fig:power-recomb} the relationship between power and local recombination rate for our sweep simulations.
    We see that power to detect sweeps is
    a decreasing function of local recombination rate for \sweepfinder and $pi$ outliers,
    which is consistent with sweeps having a larger genomic footprint in regions of low
    recombination.
    \diploshic on the other hand shows a slight increase in power with recombination rates,
    probably because it is designed to distinguish regions where a sweep happend from regions linked to a sweep.
    We also see that the presence of purifying selection acting on exons has a relatively small effect on the power to detect sweeps,    
    but one that is present across the entire range of recombinations rates observed in human chromosome 1. 
    Unsurprisingly, the empirical cutoff of $\pi$ to detect sweeps is the less robust to the effects of recombination rate than the other two methods,
    which are specifically designed to find selective sweeps.
